---
title: "Features from MIBI Segmentations"
author: Kris Sankaran^[ksankaran@wisc.edu]
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Features from MIBI Segmentations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  sample_id: 5
---

For the MIBI data, we have segmentation masks telling us where the cells are
located. The masks themselves are not interesting. However, they give
information about cell-ecosystem heterogeneity and composition which *is* very
interesting. This script produces some intermediate outputs that get more
directly at those questions,

* Polygons: We convert the raster into shapefiles, so that we can apply
  geometric logic to them (getting centroids, areas, borders, ...)
* Graphs: We can think of each cell as a node and declare edges if the cells are
  within a certain distance / are K nearest neighbors.
* Derived Statistics: For either polygons or graphs, we show how you can get
  statistics from local neighborhoods, like the proportion of the local
  neighborhood that's background.

```{r}
library("dplyr")
library("ggplot2")
library("igraph")
library("raster")
library("readr")
library("reshape2")
library("sf")
library("spdep")
library("stars")
library("stringr")
library("BIRSBIO2020.scProteomics.embeddings")
theme_set(theme_bw() + theme(panel.grid=element_blank()))
```

First, we'll download all the data, if they're not already present in the
`$HOME/Data` folder.

```{r}
# if data folder doesn't exist, download and unzip
data_paths <- download_data(file.path(Sys.getenv("HOME"), "Data"))
sample_id <- params$sample_id
f <- list.files(data_paths[[3]], str_interp("p${sample_id}_"), full.names = TRUE)
im <- raster(f)
im <- crop(im, extent(im, 0, 250, 0, 250)) # comment out to run on full data
```

# Polygonization

It can be helpful to work directly with polygon geometries, rather than the
original raster image. You can then use any logic you'd have used for
manipulating geographic shapefiles, for example.

```{r}
polys <- st_as_stars(im) %>%
  st_as_sf(merge = TRUE) %>%
  st_cast("POLYGON")
colnames(polys)[1] <- "cellLabelInImage"
```

This will join polygons that are neighbors. Notice that there is a thin border
between cells, so we'll ignore that using the `snap` argument in `poly2nb`.

```{r}
cell_data <- read_csv(file.path(data_paths[[3]], "cellData.csv")) %>%
  filter(SampleID == sample_id)

polys <- polys %>%
  inner_join(cell_data) %>%
  group_by(cellLabelInImage) %>% # some regions get split into two adjacent polys --> merge
  inner_join(cell_data) %>%
  group_by(cellLabelInImage) %>%
  summarise_all(function(x) { x[1] })
```

With these shapefiles, we can use geographic plotting tools. For example, this
plots the immune group for each geometry.

```{r}
ggplot(polys) +
  geom_sf(aes(fill=as.factor(immuneGroup))) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(legend.position = "bottom")
```

# Graph Construction

Sometimes we care less about the locations / sizes of individual cells than
their connectivity / mixing structure. In the plot below, nodes are located at
the original cell positions, and contiguity and knn edges are blue and red,
respectively.

```{r}
nb_contig <- poly2nb(polys, snap=1)
coords <- polys %>%
  .[["geometry"]] %>%
  st_centroid() %>%
  st_coordinates()

nb_knn <- knn2nb(knearneigh(coords, k = 5))
plot(polys$geometry[seq_len(nrow(polys))])
plot(nb_knn, coords, add=T, col="#476a79")
plot(nb_contig, coords, add=T, col="#97293f")
```

We can turn this into a more familiar `igraph` object, for access to the usual
graph manipulation routines.

```{r}
G <- extract_graph(polys)
plot(G, vertex.size=3, edge.arrow.size=0, vertex.label=NA)
```

# Raster Window Features

With these new data structures, we can compute some cell-level summaries. First,
let's look at features defined on windows centered around individual cells.

The function below applys `fun` to the buffered window around the cell with
label `cell_id`. For example, we can use it to get the proportion of each type
of immune group within a window around that cell. This can be used to define a
localized cell heterogeneity, for use in downstream analysis. We plot each of
the image subwindows surrounding a single cell.

```{r}
#' How dense are cells, relative to background?
cell_ids <- unique(polys$cellLabelInImage)
loop_stats(cell_ids[1:10], "raster", im, polys, background_prop)
```

# Neighbor Features

Alternatively, we could ignore the spatial information entirely, and summarize
cells according to the properties of their neighbors.

```{r}
loop_stats(cell_ids[1:10], "graph", G, polys, background_prop)
```
