---
title: "Mapping between Embeddings"
author: Kris Sankaran^[ksankaran@wisc.edu]
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping between Embeddings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  qsize: 100 # size of image quadrant we subset to
  max_rank: 2000 # rank at which we start thresholding
  cluster_K: 20 # number of clusters in the umap layout space
  max_samples: 10 # number of samples to run on
---


* In `composition.rmd` we ask whether it's possible to recover sample-level
  spatial features from protein expression measurements
* We ask a more challenging question here, looking for whether it's possible to
  identify spatial features associated with individual cells, just from the
  expression measurements 
* Understandably more far fetched, but there are some reasons it could happen --
  maybe when an immune cell is surrounded by cancer cells, it has a different
  protein expression pattern than when it is surrounded by other immune cells
* Again, motivation is that spatial characteristics of cells are interesting,
  but it's easier to use technologies that only interrogate expression
  information

One theme between this vignette and `composition.Rmd` is that the notion of a
sampling unit is not so straightforwards in this analysis -- we can reasonably
analyze the data at several levels of resolution. This vignette considers the
cell-resolution analog of the sample-level questions in that vignette.

Without further delay, these are the packages we use in this analysis.

```{r}
library("SummarizedExperiment")
library("dplyr")
library("forcats")
library("ggplot2")
library("keras")
library("raster")
library("reshape2")
library("stars")
library("stringr")
library("tibble")
library("tidyr")
library("umap")
library("viridis")
library("BIRSBIO2020.scProteomics.embeddings")
theme_set(theme_bw() + theme(panel.grid=element_blank()))
```

We download the data if it's not already present. We're filtering to just 10
samples in this run, since it lets the vignette run faster. But to train the
complete model, replace the `params$max_sample` with 41 (the number of raster
files we have).

```{r}
data_dir <- file.path(Sys.getenv("HOME"), "Data")
get(load(file.path(data_dir, "mibiSCE.rda")))
download_data(data_dir)
loaded_ <- load_mibi(data_dir, params$max_samples)
```
We'll loop over all the tiffs and extract the `r params$qsize` $\times$ `r params$qsize` 
top-left quadrant from each image, also to speed up computation in these experiments.

```{r}
subsample <- spatial_subsample(loaded_$tiffs, loaded_$mibi)
ims <- subsample$ims
mibi_sce <- subsample$exper
```

Now, let's extract some features on which to perform the joint embedding. We'll
transform and reweight the columns, to make the two sets of features more
comparable. First, for transformation, we'll convert antigen expression values
to ranks and then threshold.

```{r, message=TRUE}
x <- t(assay(mibi_sce))
x_order <- hclust(dist(t(x)))$order
pairs(x[, x_order[1:6]], col = rgb(0, 0, 0, 0.5), cex=0.1)
```

Next, we'll extract some features from the spatial data. Note that some cells
seem to appear in the raster but not in the `colData`. This seems weird, and is
worth looking into, but for now I'm going to just innerJoin to ignore that.

```{r graph_stats}
# polygonize each raster
col_df <- as.data.frame(colData(mibi_sce)) %>%
  mutate(
    cell_type = cell_type(mibi_sce),
    cell_group = fct_lump(cell_type, prop = 0.05),
  )

polys <- list()
for (i in seq_along(ims)) {
  cur_sample <- names(ims)[i]
  cur_cols <- col_df %>%
    filter(SampleID %in% cur_sample)
  polys[[i]] <- polygonize(ims[[i]]) %>%
    mutate(SampleID = as.numeric(cur_sample)) %>%
    unite(sample_by_cell, SampleID, cellLabelInImage, remove=FALSE) %>%
    inner_join(cur_cols, by = c("SampleID", "cellLabelInImage"))
}

polys <- polys[sapply(polys, nrow) > 0]
col_df <- col_df %>%
  unite(sample_by_cell, SampleID, cellLabelInImage, remove=FALSE)

# a little plot
polys_df <- do.call(rbind, polys)
ggplot(polys_df %>% filter(!is.na(cellSize))) +
  geom_sf(aes(fill = as.factor(tumorYN))) +
  facet_wrap(Survival_days_capped_2016.1.1~SampleID) +
  theme(legend.position = "bottom")

# some features that don't need graph construction
cell_stats <- polys_df %>%
  dplyr::select(sample_by_cell, cell_type, cellSize) %>%
  mutate(
    log_size = log(cellSize),
    value = 1
  ) %>%
  spread(cell_type, value, 0) %>%
  as.data.frame() %>%
  dplyr::select(-cellSize, -geometry)

# extract basic graph features
graph_stats <- list()
for (i in seq_along(polys)) {
  print(paste0("processing sample ", i, "/", length(polys)))
  cell_ids <- unique(polys[[i]]$cellLabelInImage)
  G <- extract_graph(polys[[i]])
  graph_stats[[i]] <- loop_stats(cell_ids, "graph", G, polys[[i]], type_props) %>%
    mutate(cellType = paste0("graph_neighbors_", cellType)) %>%
    spread(cellType, props, 0) %>%
    mutate(SampleID = first(polys[[i]]$SampleID))
}

graph_stats <- bind_rows(graph_stats) %>%
  mutate_all(replace_na, 0) %>%
  unite(sample_by_cell, SampleID, cellLabelInImage) %>%
  left_join(as.data.frame(cell_stats))

# example plot: what are the neighborhoods of tumors?
plot(graph_stats$`Keratin-positive tumor`, graph_stats$`graph_neighbors_Keratin-positive tumor`)
```

Now, we'll standardize these features and learn some embeddings.

```{r}
# standardize features between the two tables
x <- t(assay(mibi_sce)) %>%
  as_tibble()

x <- x %>%
  mutate_all( function(u) (u - min(u)) / diff(range(u))) %>%
  mutate_all(function(u) u / sqrt(ncol(x))) %>%
  dplyr::select_if(function(u) !all(is.na(u))) %>%
  mutate(sample_by_cell = col_df$sample_by_cell)

y <- graph_stats %>%
  mutate_at(vars(-sample_by_cell), function(u) (u - min(u)) / diff(range(u))) %>%
  mutate_at(vars(-sample_by_cell), function(u) u / sqrt(ncol(graph_stats) - 1)) %>%
  select_if(function(u) !all(is.na(u)))

z <- x %>% inner_join(y, by = "sample_by_cell")
z_mat <- z %>%
  dplyr::select(-sample_by_cell) %>%
  as.matrix()
heatmap(z_mat)

# learning embeddings across the two tables
conf <- umap.defaults
conf$min_dist <- 0.8
embeddings <- umap(z_mat, conf)
embeddings_df <- embeddings$layout %>%
  as_tibble(.name_repair = "universal") %>%
  rename("l1" = `...1`, "l2" = `...2`) %>%
  mutate(sample_by_cell = z$sample_by_cell) %>%
  left_join(col_df) %>%
  left_join(z)
```

We'll plot the embeddings we just made, against some of the derived features.

```{r}
ggplot(embeddings_df) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(
    aes(x = l1, y = l2, col = cell_group),
    size = 0.5, alpha = 0.7
  ) +
  facet_wrap(~SampleID) +
  scale_color_brewer(palette = "Set2") +
  guides(col = guide_legend(override.aes = list(alpha = 1, size = 5))) +
  ggtitle("Cell Types") +
  theme(legend.position = "bottom")

ggplot(embeddings_df) +
  geom_point(aes(x = l1, y = l2, col = Fe)) +
  facet_wrap(~ SampleID) +
  scale_color_viridis(option = "inferno")

ggplot(embeddings_df) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(
    aes(x = l1, y = l2, col = sqrt(graph_neighbors_Macrophages)),
    size = 0.5, alpha = 0.7
  ) +
  scale_color_viridis(option = "inferno") +
  facet_wrap(~ SampleID) +
  theme(legend.position = "bottom") +
  ggtitle("Prop. Neighbors are Macrophages")
```

# Cells $\to$ Samples

Now that we have embeddings at the cell level, we can try to summarize samples
by how many of their cells lie in different regions of the embedding space.
First, we'll cluster using an arbitrary $K$.

```{r}
clusters <- kmeans(embeddings$layout, centers = params$cluster_K)

# plot the clusters
ggplot(embeddings_df) +
  geom_point(
    aes(x = l1, y = l2, col = cell_group),
    size = 0.5, alpha = 0.1
  ) +
  geom_text(
    data = data.frame(clusters$centers, cluster = seq_len(nrow(clusters$centers))),
    aes(x = X1, y = X2, label = cluster), size = 5
  ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  guides(col = guide_legend(override.aes = list(alpha = 1, size = 5))) +
  facet_wrap(~SampleID) +
  ggtitle("Clusters vs. Cell Types") +
  theme(legend.position = "bottom")

# summarize the samples
embeddings_df$cluster <- as_factor(clusters$cluster)
cluster_props <- embeddings_df %>%
  group_by(SampleID, cluster) %>%
  summarise(count = n()) %>%
  group_by(SampleID) %>%
  mutate(
    total = sum(count),
    prop = count / total
    ) %>%
  left_join(
    col_df %>%
      dplyr::select(SampleID, GRADE) %>%
      unique()
  )

ggplot(cluster_props) +
  geom_bar(
    aes(x = as.factor(SampleID), y = prop, fill = cluster),
    position = "stack", stat = "identity"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  facet_grid(.~GRADE, scales="free_x", space="free_x")
```

Question: Can you map the different clusters back to the spatial patterns that
we had noticed from before? E.g., are those in cluster "5" more spatially
heterogeneous?

To answer this, let's compare samples 23, 27, 2 with 10 and 12. These samples
seem to have pretty different cluster compositions, even though they both lie in
the grade 3 group.

```{r}
polys_q <- polys_df %>%
  filter(SampleID %in% c("23", "27", "2", "10", "12")) %>%
  mutate(compare = ifelse(SampleID %in% c("10", "12"), "group2", "group1"))

ggplot(polys_q) +
  geom_sf(aes(fill = as.factor(tumorYN))) +
  facet_wrap(compare ~ SampleID) +
  theme(legend.position = "bottom")
```

I'm not sure I'd have put 10 and 12 that close to one another, since 10 has lots
of empty space, but at least the first three do seem qualitatively similar.

In fact, we can compare many of the spatial images, based on the ordering we
just defined.

```{r}
cluster_props_wide <- cluster_props %>%
  dcast(SampleID ~ cluster, value.var = "prop", fill = 0)

 sample_order <- cluster_props_wide %>%
  dist() %>%
  hclust()
sample_order <- sample_order$order

polys_df <- polys_df %>%
  mutate(SampleID = factor(SampleID, levels = as.character(cluster_props_wide$SampleID[sample_order])))

ggplot(polys_df %>% filter(SampleID %in% levels(polys_df$SampleID)[1:12])) +
  geom_sf(aes(fill = as.factor(tumorYN))) +
  facet_wrap(GRADE ~ SampleID) +
  theme(legend.position = "bottom")

tmp <- cluster_props %>%
  left_join(
    col_df %>%
      dplyr::select(SampleID, GRADE) %>%
      unique()
  )

ggplot(tmp) +
  geom_dotplot(
    aes(fill = as.factor(GRADE), x = prop)
  ) +
  facet_wrap(~ cluster)
```

# Guessing the Embedding

We'll finally load the cytof data to identify shared features. Then, we'll look
at the relationship between those features and the above embeddings, to see if
we could map the cytof data into the mibitof embedding space.

```{r}
load(file.path(data_dir, "masstagSCE.rda"))
masstag <- data_list("sce")
common_antigens <- intersect(rownames(masstag$tcell.sce), rownames(mibi_sce))
```

I'll learn a mapping at the cell level, using (1) cell identity and (2) levels
of `r common_antigens` to identify the corresponding location in the embedding
space. If this is accurate, we could then cluster the imputed cell embeddings,
to come up with a new spatial summary of the mass spec data. Otherwise, we would
report that the antigen information on its own is not enough to determine the
spatial characteristics of the sample.

This prepares the training data, which maps antigen values and cell type
indicators to embedding locations.

```{r}
predictors <- x %>%
  left_join(
    y %>%
    dplyr::select(-starts_with("graph")) %>%
    rename_all(tolower),
    by = "sample_by_cell"
  )

response <- embeddings_df %>%
  dplyr::select(l1, l2) %>%
  as.matrix()

pred_mat <- predictors %>%
  dplyr::select(all_of(common_antigens)) %>%
  as.matrix()

model <- generate_model(ncol(pred_mat))
model %>%
  fit(pred_mat, response, epochs=200, batch_size=1024)
```

```{r}
y_hat <- predict(model, pred_mat)
plot(response, cex=0.1, col = rgb(0, 0, 0, 0.1))
points(y_hat, cex=0.1, col = rgb(1, 0, 0, 0.1))
```

What happens when we summarize these samples by these predicted cluster
memberships? Ideally, you would be able to recognize the original cluster
compositions, which should reflect spatial expression patterns (if they don't
already).

In the worst case, the imputed cluster compositions would be totally unrelated
to the real (known, for the MIBI-ToF data) compositions.


We can compare this with what the predictions would have been if we had used all
the antigens (but no explicit spatial information).

```{r}
pred_mat <- predictors %>%
  dplyr::select(-sample_by_cell) %>%
  as.matrix()

model <- generate_model(ncol(pred_mat))
model %>%
  fit(pred_mat, response, epochs=200, batch_size=1024)
```

```{r}
y_hat <- predict(model, pred_mat)
plot(response, cex=0.1, col = rgb(0, 0, 0, 0.1))
points(y_hat, cex=0.1, col = rgb(1, 0, 0, 0.1))
```

[^1]: We can try predicting different sample characteristics from the
    embeddings, for example. It seems like what people do by eye anyways (trying
    to tell whether known group separate).
