---
title: "Interactive Visualization using Linked Views"
author: Kris Sankaran^[ksankaran@wisc.edu]
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive Visualization using Linked Views}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
---

This vignette highlights an approach to integrative analysis that leverages
interactive visualization. Sometimes, the relationships between data types can
be difficult to determine algorithmically, but may become clear upon direct
manipulation.

The high-level idea is that traditional, single-table analysis output something
like a map -- a visual device that let's us make comparisons. The appeal of many
dimensionality reduction or trajectory reconstruction techniques is that we can
survey the landscape of cells at a glance. The algorithms give reductions of the
raw data that are easy to navigate: isolated islands or gradually evolving
rivers of cell types are easier to recognize than if we had to inspect the raw
input counts.

Continuing this analogy, this suggests that a goal of integrative analysis is to
be able to relate pairs of maps. In the visualizations below, we use the linking
principle [@buja1991interactive] to relate the abstract U-Map constructed from
protein expression levels with the physical spatial map produced by MIBI-TOF.
Linking let's us estimate the probability that a cell lies in a particular
location in one map, conditional on its location in the other.

Since we want to be able to generate these interactive views on demand, we wrap
the core javascript visualization code in an
[htmlwidgets](https://www.htmlwidgets.org) package.

# Setup

First, we'll load packages needed to prepare the data and run the analysis. Note
that we use spatial data analysis packages to prepare the geojson which ends up
as input to the visualization

```{r}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r}
library("dplyr")
library("geojsonio")
library("raster")
library("readr")
library("rmapshaper")
library("umap")
library("BIRSBIO2020.scProteomics.embeddings")
```

Next, we load (or download) the MIBI-TOF data.

```{r}
data_dir <- file.path(Sys.getenv("HOME"), "Data")
download_data(data_dir)
loaded_ <- load_mibi(data_dir)
cell_full <- read_csv(file.path(data_dir,  "TNBC_shareCellData", "cellData.csv"))
```

Our visualizations will be made on a patient-by-patient basis. While in
principle it would be possible to construct a visualization across all patients
simultaneously, laying out the spatial maps in an intelligent way introduces
complexity. For example, it would be more informative if the cell layouts of
similar patients were placed closer to one another. For this reason, we postpone
the problem of studying patients together, and focus on the usability of linking
when working with one patient at a time.

The function below encapsulates all preprocessing that has to be performed for a
single patient. The input are the full protein expression data, paths to all
spatial Tiffs, and the ID of the patient to display. The output are a data frame
giving the U-Map coordinates of all cells, along with a geojson giving the
spatial layout of those cells. We simplify the geojson, to avoid creating heavy
computation during interaction. We also crop the rasters, so that preprocessing
(and the compilation of this vignette) is faster.

```{r}
input_data <- function(cell_full, tiff_paths, sample_id, crop_size=1000, simplify=0.4) {
  # read the associated raster
  im <- raster(tiff_paths[grepl(paste0("p", sample_id, "_"), tiff_paths)])
  im <- crop(im, extent(im, 0, crop_size, 0, crop_size)) # comment out to work on full data
  polys <- polygonize(im) %>%
    filter(!(cellLabelInImage %in% c(0, 1)))

  # filter down to relevant data
  cell_data <- cell_full %>%
    filter(
      SampleID == sample_id,
      cellLabelInImage %in% polys$cellLabelInImage
    )

  polys <- polys %>%
    left_join(cell_data)

  # run U-Map on channel data
  channels <- setdiff(colnames(cell_data), c("cellLabelInImage", "SampleID", "cellSize", "Background", "tumorYN", "cell_cluster", "tumorCluster", "Group", "immuneCluster", "immuneGroup"))
  dimred <- umap(cell_data[, channels]) %>%
    .[["layout"]] %>%
    data.frame() %>%
    rename(V1 = X1, V2 = X2)
  dimred <- cbind(cell_data, dimred)

  # convert to geojson
  geo_polys <- geojson_json(polys) %>%
    geojson_sp() %>%
    ms_simplify(keep = simplify) %>%
    geojson_json()

  list("dimred" = dimred, "polys" = geo_polys)
}
```

Below is the visualization applied to the output of this function. Each pair of
panels corresponds to the subset of cells from a region of tissue within a
single patient. On the left hand side are the physical locations of each cells,
shaded in according to cell type. On the right hand side are the abstract
coordinates described by U-Map. For completeness, the cell types are shaded
according to

* Tumor clusters: <span style="background:#d9d9d9">4</span>, <span style="background:#bc80bd">7</span>, <span style="background:#ccebc5">10</span>, <span style="background:#ffed6f">17</span>.
* Immune clusters: <span style="background:#8dd3c7">1</span>, <span style="background:#ffffb3">2</span>, <span style="background:#bebada">3</span>, <span style="background:#fb8072">4</span>, <span style="background:#80b1d3">8</span>, <span style="background:#fdb462">10</span>, <span style="background:#b3de69">11</span>, <span style="background:#fccde5">12</span>.

Click and drag over a set of cells (in either panel) to select them. The
corresponding locations of those cells in the other panel are highlighted when
you do so. If a brush is added to each panel, then all cells in the union of the
two selections are highlighted. See the example
[here](https://youtu.be/snzJ8Xa89LA) for typical use.

```{r, warning=F, message=F}
data_ <- list()
for (i in seq_len(6)) {
  data_[[i]] <- input_data(cell_full, loaded_$tiffs, i)
}

linked_views(data_[[1]]$polys, data_[[1]]$dimred)
linked_views(data_[[2]]$polys, data_[[2]]$dimred)
linked_views(data_[[3]]$polys, data_[[3]]$dimred)
linked_views(data_[[4]]$polys, data_[[4]]$dimred)
linked_views(data_[[5]]$polys, data_[[5]]$dimred)
linked_views(data_[[6]]$polys, data_[[6]]$dimred)
```
